---
title: "PM 591 Final Project"
author: "Zoe Tsokolas"
date: "2025-04-09"
output: html_document
---

## wd
```{r}

setwd("/Users/zoetsokolas/Desktop/PM591/ML_ABPos_WMH_FinalProject/")

library("mlr3")
library("mlr3learners")
library("mlr3tuning")
library("ggplot2")
library("paradox")
library("mlr3viz")
library("fastDummies")
library("dplyr")

```

- goal is to "build a predictive model using real data
- describe the problem, modeling approaches, and results
- need to use 3 distinct, appropriate methods
- describe how i split data into train/test and any resampling strategy like cross validation used for comparing models, tuning, or feature selection
- describe the model/feature selection procedures
- describe tuning parameters and my tuning approach
- describe performance metrics used and my choices (ex. AUC). 
- provide key data summaries using tables and or plots (plots preferred) (such as class frequencies if doing classification) 
- report training, validation/CV, and test errors
- include CV plots, variable importance measures (p vals, model coefficients, feature importance from tree based models)
- discuss whether the predictive models developed achieved sufficient accuracy to be useful for predicting new observations based on 1. practical utility of the model, 2. strengths and limitations, and 3. possible improvements or extensions

## data cleaning / preprocessing
```{r}

# reading in data
data <- read.csv("HD1_data_r6_cleaned_merged - HD1_data_r6_cleaned_merged.csv")

# subsetting only CU 
CUdata <- data[data$CDX_Cog == 0, ]
table(CUdata$CDX_Cog) # should have only counts for "0" 

# converting E4 status and Amyloid Positivity to single binary feature 
# because otherwise ML models interpret E4 = 0 and E4 = 1 in same person (everyone has a value of E4- or E4+ = ex. they'll have "1" for the feature APOE4_Positivity.X0 if they're E4- but ALSO have "0" encoded for feature APOE4_Positivity.X1... so ML models might "double count")
CUdata$APOE4_Positive <- ifelse(CUdata$APOE4_Positivity == "1", 1, 0)
CUdata$X01_AB_FBB_AB_pos <- ifelse(CUdata$X01_AB_FBB_AB_pos == "1", 1, 0)

# scanner had weird extra empty string of "" w/no values, removing it
CUdata$X01_AB_FBB_Scanner<-as.character(droplevels(as.factor(CUdata$X01_AB_FBB_Scanner)))

# inspecting continous data using shapiro wilk test
shapiro.test(CUdata$X02_WMH_Volume_Raw)
shapiro.test(CUdata$Age)
shapiro.test(CUdata$X01_ICV_HD)
shapiro.test(CUdata$ID_Education)
shapiro.test(CUdata$BW_HDLChol)
shapiro.test(CUdata$BW_LDLchol)
shapiro.test(CUdata$BW_Cholhdlcratio)
shapiro.test(CUdata$BW_Triglycerides)
shapiro.test(CUdata$OM_BMI)
shapiro.test(CUdata$CDX_Diabetes)
shapiro.test(CUdata$BW_HBA1c)
shapiro.test(CUdata$IMH_Diabetes)
shapiro.test(CUdata$OM_AbCircumference)

## log-transforming non-normally distributed variables (W = close to 1 in test above = pretty normal)
CUdata$X02_WMH_Volume_LOGGED <- log10(CUdata$X02_WMH_Volume_Raw + 1)

## fixing diabetes variable
CUdata <- CUdata %>% mutate(CDX_Diabetes=case_when(
  BW_HBA1c >= 6.5 | IMH_Diabetes == 1 ~ "1",
  IMH_Diabetes == 0 & BW_HBA1c < 6.5 ~ "0"
))
CUdata <- CUdata %>% mutate(CDX_Diabetes = case_when(
  BW_HBA1c >= 6.5 | IMH_Diabetes == 1 ~ "1",
  DA_Aware == 1 ~ "1",  
  IMH_Diabetes == 0 & BW_HBA1c < 6.5 ~ "0",
  TRUE ~ CDX_Diabetes
))
CUdata <- CUdata %>% filter(!is.na(CDX_Diabetes))
CUdata_check <- CUdata %>% filter(CDX_Diabetes == "0" & DA_Aware == 1)

# ensuring complete cases for final analysis dataset
variables <- c("Med_ID", "Age", "ID_Gender", "Ethnicity", "ID_Education", "ID_Hispanic", "X01_AB_FBB_AB_pos", "X01_AB_FBB_Scanner", "X02_WMH_Volume_LOGGED", "X01_ICV_HD", "APOE4_Positivity", "BW_HDLChol", "BW_LDLchol", "BW_Cholhdlcratio", "BW_Triglycerides", "OM_BMI", "CDX_Diabetes", "BW_HBA1c", "IMH_Diabetes", "CDX_Hypertension", "Smoke_Ever", "OM_AbCircumference")
final_doc <- CUdata[complete.cases(CUdata[, variables]), variables]
final_doc <- droplevels(final_doc) # dropping unused levels from factor variables
nrow(final_doc) # 804 
any(is.na(final_doc)) # should be FALSE
# nrow(CUdata) - nrow(final_doc) ## <- can uncomment to see how many got dropped

# recoding factors with 2 levels (binary) to numeric (bc ENet and XGBoost expect numeric input, not factors)
final_doc$ID_Gender <- ifelse(final_doc$ID_Gender == "Female", 1, 0) # male == 0, female == 1
final_doc$X01_AB_FBB_Scanner <- ifelse(final_doc$X01_AB_FBB_Scanner == "Vision", 1, 0) # MCT20 == 0, vision ==1
final_doc$ID_Hispanic <- ifelse(final_doc$ID_Hispanic == 0, 0, 1) ## non hispanic == 0, hispanic == 1 - 5

# ensuring numeric because for some reason they were stored as integers before
final_doc[c("BW_HDLChol", "BW_LDLchol", "BW_Cholhdlcratio", "BW_Triglycerides")] <- lapply(final_doc[c("BW_HDLChol", "BW_LDLchol", "BW_Cholhdlcratio", "BW_Triglycerides")], function(x) as.numeric(as.character(x)))

# uncomment if you want to make sure no NAs colSums(is.na(final_doc[, c("BW_HDLChol", "BW_LDLchol", "BW_Cholhdlcratio", "BW_Triglycerides")]))

### 
# ethnicity has 3 levels...
# have to encode categorical variable for linear model or else you'd get multicollinearity
# white == reference
# this adds "Ethnicity_Black" which == 1 if Black, 0 if not 
# also adds "Ethnicity_Hispanic" which == 1 if Hispanic, 0 if not 
# and Ethnicity_White is implicitly encoded so if black and hispanic == 0 then they're white

final_doc <- dummy_cols(
  final_doc,
  select_columns = "Ethnicity",
  remove_first_dummy = TRUE,      # avoids dummy variable trap
  remove_selected_columns = TRUE  # drops original 'Ethnicity' column
)

## sanity check to ensure it worked (all should be FALSE)
sapply(final_doc, is.factor)

# z-scoring continuous variables
final_doc[c("Age", "ID_Education", "X01_ICV_HD", "X02_WMH_Volume_LOGGED", "BW_HDLChol", "BW_LDLchol", "BW_Cholhdlcratio", "BW_Triglycerides", "OM_BMI", "CDX_Diabetes", "BW_HBA1c", "IMH_Diabetes", "OM_AbCircumference")] <-
  lapply(final_doc[c("Age", "ID_Education", "X01_ICV_HD", "X02_WMH_Volume_LOGGED", "BW_HDLChol", "BW_LDLchol", "BW_Cholhdlcratio", "BW_Triglycerides", "OM_BMI", "CDX_Diabetes", "BW_HBA1c", "IMH_Diabetes", "OM_AbCircumference")], function(x) as.numeric(scale(as.numeric(x))))

# removing Med_ID so it doesn't influence the model as a predictor -- identifiers shouldn't be used as feature in ML models (unless your model is grouping) 
final_doc <- final_doc[, !names(final_doc) %in% "Med_ID"]

# to check original (pre z-scored) age range, mean, SD
range(CUdata$Age[complete.cases(CUdata[, variables])])
age_vals <- CUdata$Age[complete.cases(CUdata[, variables])]
mean(age_vals)
sd(age_vals)

```

```{r}

# checking dimensionality
ncol(final_doc) - 1  # subtracts outcome column -- i have 10  

```

## setup ML pipeline
```{r}

# splitting data into train and test set
# set seed for reproducibility
set.seed(101) 
# below: creates a sequence up to total # of rows in real dataset -- randomly selecting row #s to be stored in training
train_rows <- sample(seq_len(nrow(final_doc)), size = 0.7 * nrow(final_doc))
train_data <- final_doc[train_rows, ]
test_data  <- final_doc[-train_rows, ]

# classification tasks
train_data$X01_AB_FBB_AB_pos <- as.factor(train_data$X01_AB_FBB_AB_pos)
test_data$X01_AB_FBB_AB_pos <- as.factor(test_data$X01_AB_FBB_AB_pos)

task <- TaskClassif$new(id = "amyloid", backend = train_data, target = "X01_AB_FBB_AB_pos")

# defining learners
lrn_enet   <- lrn("classif.cv_glmnet", predict_type = "prob")
lrn_rf     <- lrn("classif.ranger", predict_type = "prob")
lrn_xgb    <- lrn("classif.xgboost", predict_type = "prob")
lrn_logreg <- lrn("classif.log_reg", predict_type = "prob")

xgb_param_space <- ps(
  nrounds = p_int(50, 300),
  eta = p_dbl(0.01, 0.3),
  max_depth = p_int(2, 6)
)

# tuning my XGBoost model using inner 3 fold CV (using autotuner)
auto_xgb <- AutoTuner$new(
  learner = lrn_xgb,
  resampling = rsmp("cv", folds = 3), # inner CV
  measure = msr("classif.auc"),
  search_space = xgb_param_space,
  terminator = trm("evals", n_evals = 30),
  tuner = tnr("random_search")
)

auto_xgb$train(task)
# see best parameters for XGBoost
print(auto_xgb$tuning_result)

# outer CV benchmark now
learners <- list(lrn_enet, lrn_rf, lrn_logreg, auto_xgb)

cv10 <- rsmp("cv", folds = 10)

bmr <- benchmark(
  benchmark_grid(
    tasks = task,
    learners = learners,
    resamplings = cv10
  )
)

auc_results <- bmr$aggregate(msr("classif.auc"))
auc_table <- as.data.frame(auc_results)[, c("learner_id", "classif.auc")]

autoplot(bmr, measure = msr("classif.auc"))

```

```{r}

# ensuring factor
test_data$X01_AB_FBB_AB_pos <- as.factor(test_data$X01_AB_FBB_AB_pos)

test_task <- TaskClassif$new(id = "amyloid_test", backend = test_data, target = "X01_AB_FBB_AB_pos")
# defining final ENet model
enet_final <- lrn("classif.cv_glmnet", predict_type = "prob")
# train on full training set
enet_final$train(task)
# predict on test set
enet_pred <- enet_final$predict(test_task)
# compute AUC on test set
enet_auc <- enet_pred$score(msr("classif.auc"))

# random forest
rf_final <- lrn("classif.ranger", predict_type = "prob")
rf_final$train(task)
rf_pred <- rf_final$predict(test_task)
rf_auc <- rf_pred$score(msr("classif.auc"))

# logistic regression
logreg_final <- lrn("classif.log_reg", predict_type = "prob")
logreg_final$train(task)
logreg_pred <- logreg_final$predict(test_task)
logreg_auc <- logreg_pred$score(msr("classif.auc"))

# XGBoost (its already been tuned above)
xgb_pred <- auto_xgb$predict(test_task)
xgb_auc <- xgb_pred$score(msr("classif.auc"))

# comparing all AUCs
data.frame(Model = c("Elastic Net", "Random Forest", "Logistic Regression", "Tuned XGBoost"), Test_AUC = c(enet_auc, rf_auc, logreg_auc, xgb_auc))


```

## plot ROC curve
```{r}

# computing compute the true positive rate (tpr) and false positive rate (fpr) across various thresholds
# plot ROC and extract data

## ENET
roc_data <- autoplot(enet_pred, type = "roc", return_data = TRUE)
print(roc_data)  
# take underlying data from the ggplot
roc_df <- roc_data$data
names(roc_df)[names(roc_df) == "x"] <- "FPR"
names(roc_df)[names(roc_df) == "y"] <- "TPR"
# results
print(data.frame(n_pairs = nrow(roc_df), unique_TPRs = length(unique(roc_df$TPR)), unique_FPRs = length(unique(roc_df$FPR))))

# x-axis: false positive rate (1 – specificity)
# y-axis: true positive rate (sensitivity / recall)
# diagonal  line: random guesses (AUC = 0.5)

## RANDOM FOREST
roc_rf <- autoplot(rf_pred, type = "roc", return_data = TRUE)
print(roc_rf)

rf_df <- roc_rf$data
names(rf_df)[names(rf_df) == "x"] <- "FPR"
names(rf_df)[names(rf_df) == "y"] <- "TPR"

data.frame(Model = "Random Forest", n_pairs = nrow(rf_df), unique_TPRs = length(unique(rf_df$TPR)), unique_FPRs = length(unique(rf_df$FPR)))

## LOGISTIC REGRESSION 
roc_logreg <- autoplot(logreg_pred, type = "roc", return_data = TRUE)
print(roc_logreg)

logreg_df <- roc_logreg$data
names(logreg_df)[names(logreg_df) == "x"] <- "FPR"
names(logreg_df)[names(logreg_df) == "y"] <- "TPR"

data.frame(Model = "Logistic Regression", n_pairs = nrow(logreg_df), unique_TPRs = length(unique(logreg_df$TPR)), unique_FPRs = length(unique(logreg_df$FPR)))

## TUNED XGBOOST
roc_xgb <- autoplot(xgb_pred, type = "roc", return_data = TRUE)
print(roc_xgb)

xgb_df <- roc_xgb$data
names(xgb_df)[names(xgb_df) == "x"] <- "FPR"
names(xgb_df)[names(xgb_df) == "y"] <- "TPR"

data.frame(Model = "Tuned XGBoost",n_pairs = nrow(xgb_df),unique_TPRs = length(unique(xgb_df$TPR)),unique_FPRs = length(unique(xgb_df$FPR)))


```

## to view feature importance
```{r}

# extracting coefficients
coef_lasso <- as.matrix(coef(lasso.model$learner.model, s = "lambda.min"))
coef_ridge <- as.matrix(coef(ridge.model$learner.model, s = "lambda.min"))
coef_enet  <- as.matrix(coef(enet.model$learner.model, s = "lambda.min"))

# lasso = most selective
sort(abs(coef_lasso[,1]), decreasing = TRUE)
# ridge = keeps everything but shrinks weak features
sort(abs(coef_ridge[,1]), decreasing = TRUE)
# eNet is kind of in the middle of the other 2
sort(abs(coef_enet[,1]), decreasing = TRUE)


```

```{r}


```

